Multi-scale assessment of a grassland productivity model

### 2020-06-16 manuscript/manuscript.Rmd
final small updates before submission

abstract: |
  Grasslands provide a large amount of ecosystem services globally and forecasting grassland productivity in the coming decades will provide valuable information to land managers. Productivity models can be well-calibrated at local scales, but generally have some maximum spatial extent in which they perform well. Here we evaluate a grassland productivity model to find the optimal spatial extent for parameterization, and also for subsequently applying it in future forecasts for North America. We also evaluated the model on new vegetation types to ascertain its potential generality. We find the model most suitable when incorporating only grasslands, as opposed to also including agriculture and shrublands, and only in the Great Plains and Eastern Temperate Forest ecoregions of North America. The model was not well suited to grasslands in North American Deserts or Northwest Forest ecoregions. It also performed poorly in agriculture vegetation, likely due management activities, and shrubland vegetation, likely because the model lacks representation of deep water pools. This work allows us to perform long-term forecasts in areas where model performance has been verified, with gaps filled in by future modelling efforts.

@@

Here we evaluate a productivity model with the intention of it driving long-term forecasts. The PhenoGrass model developed by @hufkens2016 is a pulse-response productivity model with temperature and precipitation as the primary drivers. The model is parameterized using observations from the Phenocam network, which have a small resolution  (footprints of < 1ha), sub-daily sampling and sites across all major biomes. These attributes make the PhenoGrass model potentially widely applicable. We expand on the evaluation of the original study by using 445 additional site-years of data from 77 new sites. We test the model’s performance across varying combinations of North American ecoregions and vegetation types to find an optimal spatial extent and/or ecoregion for which to parameterize and apply the model. Finally we address where the model performs poorly and how productivity forecasts for these areas could be implemented or improved.

@@

The PhenoGrass model is an ecohydrology model which has interacting state variables for soil water, plant available water, and plant fractional cover [@hufkens2016]. Model inputs are daily precipitation, temperature, potential evapotranspiration (derived from the Hargreaves equation [@hargreaves1985]), and solar radiation. The primary output is fractional vegetation cover. The original model form, derived in Choler et 2010 and Choler et al. 2011, used only temperature and potential evapotranspiration and was parameterized using satellite-derived NDVI data. @hufkens2016 expanded on the original Choler model by incorporating growth and sensece restraints from temperature and solar radiation, and also included a scaling factor to convert Phenocam GCC data to a fractional cover estimate. @hufkens2016 evaluated the PhenoGrass model using 14 grassland Phenocam sites across Western North America with a total of 34 site years. They found the modelled fractional cover correlated well with annual productivity at both a daily and annual timescale.

@@

The Phenocam network is a global network of fixed, near-surface cameras capturing true-color images of vegetation throughout the day [@richardson2018a]. Using a ratio of the three RGB bands a greenness metric (GCC) is calculated from each image, resulting in a daily scale time series of canopy greeness. GCC is a unitless metric which is highly correlated with satellite derived NDVI [@richardson2018b] and flux tower derived primary productivity [@yan2019; @toomey2015]. Each Phenocam image is subset to one to several different plant vegetation types based on the field of view. These regions of interest (ROI) serve as the basis for the GCC calculation and subsequent post-processing [@seyednasrollah2019].

We downloaded all Phenocam data with ROIs of the grasslands (GR), shrublands (SH), and agricultural (AG) vegetation types for the years 2012 to 2018, totalling 91 sites and 479 site-years (Fig. 1). As input to the PhenoGrass model we used the 3-day smoothed GCC scaled, for each ROI, from 0-1. In the model parameterization each ROI time series is further transformed to a fractional cover estimate using the local mean annual precipitation (MAP) combined with a scaling factor [@hufkens2016; @donohue2013].

@@

We used a three-step process to evaluate the PhenoGrass model at different spatial extents. First, we parameterized a suite of 15 PhenoGrass models comprising four different extents to find the most appropriate one to make forecasts (Fig. 1). The largest extent used all Phenocam locations described above (84 sites). Next were all sites, respectively, within the three vegetation types (grasslands, shrublands, and agricultural), and all sites, respectively, within each of the four ecoregions (North American Deserts, Great Plains, NW Forests, Eastern Temperate Forests) [@omernik2014]. Finally, we parameterized models for each vegetation type within each ecoregion (eg. All grassland locations within the Great Plains ecoregion). All sets of parameterized models were limited to have at least five sites.

Second, using the same combinations of sites described above, we applied each model parameterization to its respective extent as well as each one below it. For example, sites within the grassland vegetation type were tested using the grassland only model as well as the all site  model. All sites within the Great Plains ecoregion were tested using its own model as well as the all site model. The shrubland vegetation sites within the N.A. Deserts ecoregion were tested using its respective model, the N.A. Deserts, and the all site model. There was no cross-validation using out of sample data in this step as it would have been computationally expensive. Rather, these fitting data metrics were treated as a best case scenario in what each model parameterization can achieve. From these results we used a threshold to select which model to evaluate further using cross-validation at the appropriate scale. For the error metric we chose the coefficient of determination calculated for each site, and averaged across all sites in the respective evaluation. We used a threshold of 0.65, which is viewed as "acceptable" for time-series models [@ritter2013].

@@

Third, models which exceed the threshold described above were subject to further evaluation. For each model we performed a leave one out cross-validation, where the model was re-fit with one Phenocam site not included in the training data, and then evaluated against the left out site. In this step a scaling coefficient to link mean annual precipitation with PhenoCam GCC was held constant at the value obtained in step two. The resulting coefficient of determination is the average among all modelled sites using their respective out of sample test.

@@

At the largest extent, where the phenograss model was parameterized using all 91 sites, the model performed poorly with an average R\textsuperscript{2} value among sites of 0.33. The error of this largest extent model when applied across different subsets of sites (Table 1, leftmost column) shows that agricultural and North American desert sites contribute most to this low R\textsuperscript{2}, while some grassland sites performed well in this regard. Models built at the ecoregion or vegetation type extent also performed poorly overall. The best model performance was achieved when models were built using a single vegetation type subset to ecoregions.

@@

```{r, out.width = "12cm", echo = FALSE, fig.cap = "Smoothed time series for all 14 grassland sites used in Hufkens et al 2016 (A) and 24 additional grassland sites added in the current study (B). Each line represents the long-term average Green Chromatic Coordinate (GCC) of a single site across all available years, smoothed using a GAM model."}
knitr::include_graphics("figs/fig3_discussion_avg_site_gcc.png")
```

>>>>

abstract: |
  Grasslands provide a large amount of ecosystem services globally and forecasting grassland productivity in the coming decades will provide valuable information to land managers. Productivity models can be well-calibrated at local scales, but generally have some maximum spatial extent in which they perform well. Here we evaluate a grassland productivity model to find the optimal spatial extent for parameterization, and thus for subsequently applying it in future forecasts for North America. We also evaluated the model on new vegetation types to ascertain its potential generality. We find the model most suitable when incorporating only grasslands, as opposed to also including agriculture and shrublands, and only in the Great Plains and Eastern Temperate Forest ecoregions of North America. The model was not well suited to grasslands in North American Deserts or Northwest Forest ecoregions. It also performed poorly in agriculture vegetation, likely due to management activities, and shrubland vegetation, likely because the model lacks representation of deep water pools. This work allows us to perform long-term forecasts in areas where model performance has been verified, with gaps filled in by future modelling efforts.

@@

Here we evaluate a productivity model with the intention of it driving long-term forecasts. The PhenoGrass model developed by @hufkens2016 is a pulse-response productivity model with temperature and precipitation as the primary drivers. The model is parameterized using observations from the Phenocam network, which have a small resolution  (footprints of < 1ha), sub-daily sampling and sites across all major biomes. These attributes make the PhenoGrass model potentially widely applicable. We expand on the evaluation of the original study by using 445 additional site-years of data from 77 new sites. We test the model’s performance across varying combinations of North American ecoregions and vegetation types to find an optimal spatial extent in which to parameterize and apply the model. Finally we address where the model performs poorly and how productivity forecasts for these areas could be implemented or improved.

@@

The PhenoGrass model is an ecohydrology model which has interacting state variables for soil water, plant available water, and plant fractional cover [@hufkens2016]. Model inputs are daily precipitation, temperature, potential evapotranspiration (derived from the Hargreaves equation, @hargreaves1985), and solar radiation. The primary output is fractional vegetation cover. The original model form, derived in Choler et 2010 and Choler et al. 2011, used only temperature and potential evapotranspiration and was parameterized using satellite-derived NDVI data. @hufkens2016 expanded on the original Choler model by incorporating growth and senescence restraints from temperature and solar radiation, and also included a scaling factor to convert Phenocam G\textsubscript{cc} data to a fractional cover estimate. @hufkens2016 evaluated the PhenoGrass model using 14 grassland Phenocam sites across Western North America with a total of 34 site years. They found the modelled fractional cover correlated well with annual productivity at both a daily and annual timescale.

@@

The Phenocam network is a global network of fixed, near-surface cameras capturing true-color images of vegetation throughout the day [@richardson2018a]. Using a ratio of the three RGB bands a greenness metric (green chromatic coordinate, G\textsubscript{cc}) is calculated from each image, resulting in a daily scale time series of canopy greeness. G\textsubscript{cc} is a unitless metric which is highly correlated with satellite derived NDVI [@richardson2018b] and flux tower derived primary productivity [@yan2019; @toomey2015]. Each Phenocam image is subset to one to several different plant vegetation types based on the field of view. These regions of interest (ROI) serve as the basis for the G\textsubscript{cc} calculation and subsequent post-processing [@seyednasrollah2019].

We downloaded all Phenocam data with ROIs of the grasslands (GR), shrublands (SH), and agricultural (AG) vegetation types for the years 2012 to 2018, totalling 91 sites and 479 site-years (Fig. 1, Table A1). As input to the PhenoGrass model we used the 3-day smoothed G\textsubscript{cc} scaled, for each ROI, from 0-1. In the model parameterization each ROI time series is further transformed to a fractional cover estimate using the local mean annual precipitation (MAP) combined with a scaling factor [@hufkens2016; @donohue2013].

@@

We used a three-step process to evaluate the PhenoGrass model at different spatial extents. First, we parameterized a suite of 15 PhenoGrass models comprising four different extents to find the most appropriate one to make forecasts (Fig. 2). The largest extent used all Phenocam locations described above (84 sites). Next were all sites, respectively, within the three vegetation types (grasslands, shrublands, and agricultural), and all sites, respectively, within each of the four ecoregions (North American Deserts, Great Plains, NW Forests, Eastern Temperate Forests) [@omernik2014]. Finally, we parameterized models for each vegetation type within each ecoregion (eg. All grassland locations within the Great Plains ecoregion). All sets of parameterized models were limited to have at least five sites.

Second, using the same combinations of sites described above, we applied each model parameterization to its respective extent as well as each one below it. For example, sites within the grassland vegetation type were tested using the grassland only model as well as the all site  model. All sites within the Great Plains ecoregion were tested using the Great Plains ecoregion model as well as the all site model. The shrubland vegetation sites within the N.A. Deserts ecoregion were tested using its respective model, the N.A. Deserts ecoregion model, and the all site model. There was no cross-validation using out of sample data in this step as it would have been computationally expensive. Rather, error metrics from these in-sample tests were treated as a best case scenario in what each model parameterization can achieve. From these results we used a threshold to select which model to evaluate further using cross-validation at the appropriate scale. For the error metric we chose the coefficient of determination (R\textsuperscript{2}) calculated for each site, and averaged across all sites in the respective evaluation. We used a threshold of 0.65, which is viewed as "acceptable" for time-series models [@ritter2013].

@@

Third, models which exceed the threshold described above were subject to further evaluation. For each model we performed a leave one out cross-validation, where the model was re-fit with one Phenocam site not included in the training data, and then evaluated against the left out site. In this step a scaling coefficient to link mean annual precipitation with PhenoCam G\textsubscript{cc} was held constant at the value obtained in step two. The resulting coefficient of determination is the average among all modelled sites using their respective out of sample test.

@@

At the largest extent, where the phenograss model was parameterized using all 91 sites, the model performed poorly with an average R\textsuperscript{2} value among sites of 0.33. The error of this largest extent model when applied across different subsets of sites (Table 1, leftmost column) shows that agricultural and North American desert sites contribute most to this low R\textsuperscript{2}, while some grassland sites performed well in this regard. Models built at the ecoregion or vegetation type extent also performed poorly overall. The best model performance was achieved when models were built using a single vegetation type subset to a single ecoregion.

@@

```{r, out.width = "12cm", echo = FALSE, fig.cap = "Smoothed time series for all 14 grassland sites used in Hufkens et al 2016 (A) and 24 additional grassland sites added in the current study (B). Each line represents the long-term average Green Chromatic Coordinate of a single site across all available years, smoothed using a GAM model."}
knitr::include_graphics("figs/fig3_discussion_avg_site_gcc.png")
```
